A lottery model of density-dependent selection

### 2017-10-15 manuscript.tex
latest

The density-dependent selection literature contains some of the earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}. This work is most famously associated with MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.

Starting with Kostitzin's

Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.

Yet selection does not only change relative genotypic proportions. Ultimately, it involves the births and deaths of individuals, and is therefore intimately connected to changes in absolute abundance \cite{evolution_demography_stuff}. This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection). As a result,

One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time. This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.

 Nevertheless, much of the density-dependent selection literature has adopted variants of the logistic model, often in the setting of Mendelian inheritance \cite{roughgarden etc}.

with genotype-specific $K$ values parameterizing growth ability at high density. A closely related approach is to add Lotka-Volterra-type inter-genotypic competition terms with $\alpha$ competition coefficients, which adds ``$\alpha$-selection'' to the mix \citealt{gill_1974,case_1974,joshi_2001}.

 variant that allows for inter-genotypic versus intra-genotypic

or  explicit resource consumption \citep{tilman_1982}, are much more complicated.

$r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.

>>>>

MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at higher densities \citep{macarthur_1962,macarthur_1967}. The $r$/$K$ scheme is mostly known, and critiqued, for its role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}. The $r$/$K$ scheme also initiated a sizeable literature on density-dependent selection  (note that Kostitzin presented similar ideas much earlier \citep{kostitzin_1939,christiansen_2004}). This literature is remarkable in that it contains some of earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}, yet it has left a complex legacy.

Although $r$ and $K$ are named after the parameters of the logistic model, $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}. Yet the ability to keep growing at ever higher densities is only one aspect of selection at high population density. Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size. This led to the notion of ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equations \citep{gill_1974,case_1974,joshi_2001}.

In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).

It is thus broadly misleading to view $r$- and $K$-selection as two main contrasting forms of selection, and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].

The true contrast hinted at by the $r$/$K$ scheme is between interaction-dependent, and interaction-independent selection. That is, selective shifts in frequency are a result of differences in growth rates between genotypes, and genotypic growth rates may differ in the absence of interactions between individuals, or as a result of interactions between individuals.

Population density is a key factor controling whether individuals interact, thereby setting the relative contributions of these forms of selection. Since a huge class of growth-affecting interactions are competitive, it seems clear that the study of density-dependent selection should be grounded a reasonable representation of competition and how its effects vary with density.

%including considerable effort devoted to showing that evolution under density-dependent selection optimizes population size in some sense, akin to a maximization of fitness.
